Continous time Markov chain models in phylogenetics and phylogeography; model extension, simulation and visualisation

### 2014-05-23 PhDThesis.tex
starting to work on computational burden

%%%%%%%%%%%%%%%%
%---PACKAGES---%
%%%%%%%%%%%%%%%%
\usepackage{lgrind}
\usepackage{cmap}
\usepackage[T1]{fontenc}
\usepackage{alltt}
\usepackage{fullpage}
\usepackage{multirow}
\usepackage[english]{babel}
\usepackage{color}
\usepackage{colortbl}
\usepackage{natbib}
\usepackage{float}
\usepackage{url}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage[all]{xy}
% \usepackage{multicol}
% \usepackage{cite}
\usepackage{booktabs}
\usepackage{framed}
\usepackage{dcolumn}
\usepackage[T1]{fontenc}
\usepackage{rotating}
\usepackage{caption}

@@

% \begin{figure}[H]
% \begin{center}
% \includegraphics[scale=0.5]{samplepath}
% \end{center}
% \caption{{ \footnotesize {\bf Sample path of the Poisson process with intensity $r=1$.} 
% }}
% \label{fig:poisson}
% \end{figure}
\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{poisson}
\end{center}
\caption{{ \footnotesize {\bf Independent realisations of the Poisson process with intensity $r=1$.}
Times between mutations are exponentially distributed.
}}
\label{fig:poisson}
\end{figure}

@@

\subsection{Computing the matrix exponent \label{seq:exponentiation}}

@@

\subsection{Continuous-time Markov chain substitution models}
% TODO markov property is inherited from Poisson process
Drawing realizations with probabilities defined by $\mathbf{P}$  gives rise to a continuous-time stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$ it holds that:

@@

%---SUBSTITUTION MODEL---%
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{figures/substitution} 
\end{center}
\caption{{ \footnotesize {\bf  Nucleotide substitution model.} 
% Markov substitution models provide the cornerstone of computational phylogenetics.
}}
\label{fig:substitution}
\end{figure}

@@

@@

\subsection{Sequence evolution on a tree}

We now move onto describing the connection between already discussed CTMC models of character sequence evolution and the phylogenetic trees.
Like before we focus on a single site of the alignment, and due to the assumption of 
% markov chain unfolding on a tree
% random walk on a graph
% simulation
% 39 ; 13

\subsection{Maximum likelihood of a sequence on a tree \label{seq:likelihood}}

% to calculat elikelihood is to reverse the evolution as it happened

@@

>>>>

%%%%%%%%%%%%%%%%
%---PACKAGES---%
%%%%%%%%%%%%%%%%
\usepackage{lgrind}
\usepackage{cmap}
\usepackage[T1]{fontenc}
\usepackage{alltt}
\usepackage{fullpage}
\usepackage{multirow}
\usepackage[english]{babel}
\usepackage{color}
\usepackage{colortbl}
\usepackage{natbib}
\usepackage{float}
\usepackage{url}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage[all]{xy}
% \usepackage{multicol}
% \usepackage{cite}
\usepackage{booktabs}
\usepackage{framed}
\usepackage{dcolumn}
\usepackage[T1]{fontenc}
\usepackage{rotating}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

% \begin{figure}[H]
% \begin{center}
% \includegraphics[scale=0.5]{samplepath}
% \end{center}
% \caption{{ \footnotesize {\bf Sample path of the Poisson process with intensity $r=1$.} 
% }}
% \label{fig:poisson}
% \end{figure}
\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{poisson}
\end{center}
\caption{{ \footnotesize {\bf Independent realisations of the Poisson process with intensity $r=1$.}
Times between mutations (the branch lengths) are \emph{iid} exponentially distributed random variables with mean $1/r$.
}}
\label{fig:poisson}
\end{figure}

@@

\subsection{Computing the matrix exponent \label{sub:exponentiation}}

@@

\subsection{Markov chain models of sequence substitution}
% TODO markov property is inherited from Poisson process
Drawing realizations with probabilities defined by $\mathbf{P}$  gives rise to a continuous-time stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$ it holds that:

@@

%---SUBSTITUTION MODEL---%
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{substitution} 
\end{center}
\caption{{ \footnotesize {\bf  Nucleotide substitution model.} 
% Markov substitution models provide the cornerstone of computational phylogenetics.
}}
\label{fig:substitution}
\end{figure}

@@

%TODO: GTR as example? see M USYB 4

@@

\subsection{Sequence evolution on a tree\label{sub:evolutionOnTree}}

We now move onto describing the connection between already discussed CTMC models of character sequence evolution and the phylogenetic trees.
Like before we focus on a single site of the alignment, and due to the assumption of independence the probability of an alignment is simply the product of probabilities calculated at individual sites.

To gain a better understanding of the process shaping the data observed at the tips of a phylogenetic tree let us consider the probelm of simulating a sequence evolution on a tree.
We start with a first character $i$ sampled at the root of the tree $x_0$ from the stationary distribution $\mathbf{\Pi}$.
Let us assume that two directly neighboring nodes are $x_j$ and $x_k$.
The character sampled at the root waits an Exponentialy distributed amount of time $t\geq0$ (see Figure \ref{fig:poisson}) before moving down the tree to one of the children nodes $x_i$.
Such traversal is said to be following the pre-order, i.e. parental nodes are visited before child nodes.   
The process then randomly decides by a coin flip to which of the two neighbouring nodes to move next.
Let us assume that the process moved to $x_j$.
The state $j$ for the node $x_j$ is sampled conditional on the value $i$ already sampled at the root using the branch length expressed in real time $t$ and the vector probabilities in the row $i$ of the finite-time probabilities matrix $\mathbf{P}(t)$, calculated using SVD of the rate matrix $\mathbf{Q}$ that characterizes the CTMC.

\begin{algorithm}
\begin{center}
\begin{algorithmic}[1]
\footnotesize{
%
\State $node \gets getRoot\left(\right)$
%
\State $state \gets getState\left(node\right)$
%
\Repeat
%
\If{$\left(hasChildren\left(node\right)\right)$}
%
\ForAll{children}
%
\State $t \gets getDistanceToParent\left(child\right);$
%
\State $ \mathbf{P}\left(t\right) \gets e^{\mathbf{Q}t};$
%
\State $state \gets sample\left(\mathbf{P}\left[ state, \right]\right);$
%
\EndFor
%
\Else \Comment{this is tip node}
%
\State $node \gets getParent\left(node\right);$
%
\EndIf
%
\Until{$\left(\text{simulated for all nodes}\right)$}
}
\end{algorithmic}
\end{center}
\caption{{ \footnotesize {\bf Fabrication evolution along a phylogeny.} 
For a visited child node it's state is sampled with conditional probabilities of changing to state $j$ given state $i$ at the parental node.
}}
\end{algorithm}

This constitutes the Markov property of a CTMC, as defined in (\ref{eq:markov}).
The sampling is then repeated on the node $x_j$, before recursively proceeding to the other trio of nodes.
The states sampled at the tips of the tree constitute the character sequence for one site of the alignment.
In thi slight the CTMC process can be viewed as a continuous-time random-walk, unfolding on the topology $\mathbf{F}$, with it's behaviour described by a rate-matrix $\mathbf{Q}$.

\section{Likelihood of sequence evolution}

\subsection{Calculation of the Likelihood on a tree\label{seq:likelihood}}

Likelihood is defined as a conditional probability of observing the data given the parameters, in a function of those parameters.
Discrete molecular sequence data comes in a form of aligned matrix $\mathbf{X}$ of size $n \times l$, where $x_{jh}$ is the $h$-th character in $j$-th sequence.
Single column $\mathbf{x}_{h}$ of data matrix constitutes a single site.
Like before the standard assumption of independence posits that different sites evolve independently and that once two branches split at a node the evolution on those is also independent.

In Subsection \ref{sub:evolutionOnTree} we described how some process $\mathbf{Q}$ gives rise to the observed data $\mathbf{X}$.
In the typicall situation we start with just the observed data and aim the inference efforts at modelling the process that generated the alignment.
We can therefore denote the likelihood simply as:

\begin{equation}
L\left(\mathbf{\theta}|\mathbf{X}\right)=P\left(\mathbf{\theta}|\mathbf{X}\right), 
\label{eq:likelihood}
\end{equation}

\noindent
where $\theta$ it the topology, branch lengths, parameters of the substitution model and other parameters of interest that we aim at estimating.
This quantity expresses how likely it is to observe the data given a specific value of the parameter of interest $\theta$.
By finding the value of $\theta$ that maximizes (\ref{eq:likelihood}) we capture a model with the best possible fit.

In this thesis we will not discuss the maximum likelihood estimation itself, rather focus on closely related methods of Bayesian inference.
The likelihood is nontheless required for both methods of inference, hence we will now discuss how to calculate the likelihood of observing a certain realised sequence of characters on a tree, given a substitution model.

%---LIKELIHOOD---%
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{likelihood} 
\end{center}
\caption{{ \footnotesize {\bf  A tree with five taxa.} 
Character sequence data for single site is observed at the tips of the tree.
}}
\label{fig:LIKELIHOOD}
\end{figure}

To calculate the likelihood means thus to reverse-engineer the evolution as it happened.
Figure \ref{fig:LIKELIHOOD} presents single site $TCACC$ observed at the tips of the tree.
External nodes are numbered $x_{1}, \ldots x_{5}$, internal nodes $x_{6}, \ldots x_{8}$ and the root is placed at the node $x_{0}$.
Length of each branches leading to node $i$ is denoted as $t_{i}$.
Using methods presented in Subsection \ref{sub:exponentiation} we calculate the finite-time transition probabilities $p_{ij}(t_{i})$ between all pairs of characters in the alignment.
According to the independence assumptions to calculate the probability at a single site we need to sum over all possible combinations that could give rise to the observed sequence. Thus:

\begin{equation}
\underset{X_{0}\in S}{\sum}\underset{X_{6}\in S}{\sum}\underset{X_{7}\in S}{\sum}\underset{X_{8}\in S}{\sum}[\pi_{x_{0}}\cdot p_{x_{0}x_{8}}(t_{7})\cdot p_{x_{0}x_{7}}(t_{8})\cdot p_{x_{8}x_{6}}(t_{6})\cdot p_{x_{8}A}(t_{3})\cdot p_{x_{7}C}(t_{4})\cdot p_{x_{7}C}(t_{5})\cdot p_{x_{6}T}(t_{1})\cdot p_{x_{6}C}(t_{2})]
\label{eq:likelihood}
\end{equation}

We can break down the terms in (\ref{eq:likelihood}) to the probability of observing the data $TCACC$ for the tips and $x_{0},x_{6},x_{7},x_{8}$ for the ancestral nodes (inside the brackets) and the sum over all possible combination over the state space $S$ (outside the brackets).
Term $\pi_{0}$ is simply the probability that the root has character $x_{0}$ and comes from the assumption that the process generating the data has reached its equilibrium.

Summing over all possible paths leading to the observed data is computationally expensive, as there are  $K^{n-1}$ possible combinations, where $n-1$ is the number of internal nodes of a rooted binary tree.
Fortunately we can used the assumption of independence of the process in the two sub-trees below every node.
\citet{Felsenstein1981} provides an efficient algorithm for computing the likelihood of observing the data on a topology $\mathbf{F}$ given the substitution model that quantifies transition probabilities.
This routine, known as \emph{tree pruning} conveniently expresses the partial likelihood $L_{i}(x_{i})$ of observing data at the descendant tips of node $i$ given state $x_{i}$ at node $i$ in terms of partial likelihoods at nodes $j$ and $k$.
For all internal nodes of $\mathbf{F}$ we have:

\begin{equation}
L_{i}(x_{i})=\left[\underset{x_{j}\in \mathcal{E}}{\sum}\mathbf{P}_{x_{i}x_{j}}(t_{j})L_{j}(x_{j})\right]\times\left[\underset{x_{k}\in \mathcal{E}}{\sum}\mathbf{P}_{x_{i}x_{k}}(t_{k})L_{k}(x_{k})\right]
\label{eq:felsenstein1}
\end{equation}

\noindent
For tip nodes we have:

\begin{equation}
L_{i}(x_{i})=\begin{cases}
1 & \text{if }\hat{x}(i)=x_{i}\\
0 & \text{otherwise}
\end{cases},
\label{eq:felsenstein2}
\end{equation}

\noindent
where $\hat{x}(i)$ denotes the character state observed at node $i$.
After recursively applying (\ref{eq:felsenstein1}) to all the nodes of $\mathbf{F}$ in post-order fashion the likelihood of the data at the site is given by:
% we can integrate out the unobserved states calculating successive contributions to the partial likelihood.

\begin{equation}
L(\mathbf{x}_{h})=\underset{x_{0}\in\mathcal{E}}{\sum}\pi_{x_{0}}\cdot L_{0}(x_{0})
\label{eq:felsenstein3}
\end{equation}

\noindent
The log-likelihood of the whole alignment $X$ is computed by summing the logarithms of the likelihoods at the particular sites, due to the assumption of their independence:

\begin{equation}
l(X)=\underset{j=1}{\overset{l}{\sum}}log\left(L(\mathbf{x}_{jh})\right)
\label{eq:loglikelihood}
\end{equation}

@@

% define o notation

Calculating the log-likelihood of the tree involves:

\begin{enumerate}
\item { Singular value decomposition of the rate matrix $\mathbf{Q}$. }
\item { Taking the exponential of $\mathbf{Q}t$ for each branch length $t$. }
\item { Applying (\ref{eq:felsenstein1}) for every site in the alignment and every possible state in the state space $\mathcal{E}$. }
\item { Taking the logarithm and summing over all sites in the alignment. }
\end{enumerate}

Step (1) proceeds at ${\cal{O}}(K^3)$, however for typical, time-homogeneous applications this is not the a rate-limiting step, as repeated evaluations of $exp(\mathbf{Q}t)$ in (2) are based on the same decomposition of the matrix \citep{Suchard2009}. 
Step (2) typically boils down to repeated matrix-vector-matrix multiplications and its serial computational order is ${\cal{O}}(K^3 \times n)$. 
Step (3) takes ${\cal{O}}(K^2 \times n \times l)$ time and is the most computationally expensive part of the routine. 
Step (4) proceeds at ${\cal{O}}(l)$, linearly with the number of sites.
Fortunately these operations are very regular and lend themselves to the computational parallelism, such as utilized in the BEAGLE library \citep{Suchard2009,Ayres2012}.

% I skipped site rate variability, but perhaps it should be mentioned?
